### 2017-12-12 submission/manuscript.Rmd
update first and second result sections

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened. For feces and not tissue a decrease in evenness and Shannon diversity was associated with carcinoma [Figure 1]. Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). We next tested whether these detectable differences in community resulted in significant increases in RR. For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))` and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

@@

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
Based on the small increase in RR using $\alpha$-diversity metrics, we wanted to know if the presence of specific genera resulted in a higher RR for both stool and tissue. To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for and increase in RR. The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.

>>>>

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened. Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1]. We next tested whether these detectable differences in community resulted in significant increases in RR. For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))` and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

@@

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if the presence of specific genera resulted in a higher RR for both stool and tissue. To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for an increase in RR. The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual. In fecal samples, the RR for carcinoma due to the prescence of CRC-associated genera was greater than either the RR associated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].

### 2017-12-12 submission/manuscript.tex
update first and second result sections

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. For feces and not tissue a decrease in
evenness and Shannon diversity was associated with carcinoma {[}Figure
1{]}. Using linear mixed-effect models to control for study, resampling
of the same individual (tissue studies only), and variable region, there
was a significant decrease from control to adenoma to carcinoma for both
evenness and Shannon diversity in stool (P-value = 0.025 and 0.043,
respectively) but not for tissue (P-value \textgreater{} 0.05). We next
tested whether these detectable differences in community resulted in
significant increases in RR. For fecal samples, a decrease in evenness
resulted in a significantly increased RR for carcinoma (RR = 1.36 (1.15
- 1.61), P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value
= 0.032) while a decrease in Shannon diversity only increased the RR for
carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
Interestingly, for both adenoma and carcinoma there was no increase in
RR within tissue samples for any alpha diversity metric investigated
{[}Table S1-S3{]}.

@@

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we wanted to know if the presence of specific genera resulted
in a higher RR for both stool and tissue. To investigate this we
analyzed the classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for and increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC
associated genera was greater than either the RR assoicated with
evenness or Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR
of carcinoma continuously increased as individuals tested positive for
more CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
greater for stool (RR range = 1.62 - 2.37) than for tissue (RR range =
1.21 - 1.81). This decrease may be explained by the fact that the tissue
analysis included matched samples from the same individual.

>>>>

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using linear mixed-effect models to
control for study, resampling of the same individual (tissue studies
only), and variable region, there was a significant decrease from
control to adenoma to carcinoma for both evenness and Shannon diversity
in stool (P-value = 0.025 and 0.043, respectively) but not for tissue
(P-value \textgreater{} 0.05) {[}Figure 1{]}. We next tested whether
these detectable differences in community resulted in significant
increases in RR. For fecal samples, a decrease in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032)
while a decrease in Shannon diversity only increased the RR for
carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
Interestingly, for both adenoma and carcinoma there was no increase in
RR within tissue samples for any alpha diversity metric investigated
{[}Table S1-S3{]}.

@@

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
for an increase in RR. The majority of CRC-associated genera for both
feces and tissue had a significantly increased RR for carcinoma but not
for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
decrease may be explained by the fact that the tissue analysis included
matched samples from the same individual. In fecal samples, the RR for
carcinoma due to the prescence of CRC-associated genera was greater than
either the RR associated with evenness or Shannon diversity {[}Figure 2
\& 3{]}. Additionally, the RR of carcinoma continuously increased as
individuals tested positive for more CRC-associated genera {[}Figure 3B
\& 3D{]}.
